Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-25 paper/paper.tex
working on technique section

\section*{Problem}

The goal is to calculate a modified similiary function that leverages the number
of commonn users to better estimate the true similarity. Formally, we design
$ModSim$ such that for two items $A$ and $B$ with $n$ users in common

\begin{align}
ModSim(A, B, n) \approx TrueSim(A, B)
\end{align}
where $TrueSim$ is the true similarity of the two items. Of course, there is no
way to actually know the true similarity of two items, but for a good enough
similarity function, such as $PearsSim$ or $CosSim$ and a large enough $n$, we
can get a good estimate. Thus, we model $TrueSim$ as 
\begin{align}
TrueSim(A, B) = lim_{n\to\infty}Sim(A, B)
\end{align}
where $Sim$ is the most appropriate generic similarity function for the dataset.

Since $TrueSim$ dependends on the choice of $Sim$, we abstract away the details
of $Sim$ and use its output directly. In all, the goal is to construct $ModSim$
such that

@@

In our first approach, we model true and observed similarity probabilistically. For
this analysis, we can consider an abstact $TrueSim$ which ignores the details of 
$TrueSimRand$ or $TrueSimPref$. Similarly, we consider $Sim_{n}$ and $GoldSim$ to be
the generic observed similarity scores. Let $Y$ be a random variable representing 
the $TrueSim$ of a randomly chosen pair of items. Let $X_{n}$ be a random variable 
representing the $Sim_{n}$ of a pair of items when $n$ common users are observed.

@@

The model suggests that there is a linear correlation between $X_n$ ($Sim_n$)
and $Y$ ($TrueSim$). Moreover, the best predicted $TrueSim$ is a linear 
combination of the mean and the observed similiarity. Since $\sigma_{err,n}^2$ is 
nonnegative, the slope is always less than or equal to one, meaning that the 
observed distance from $\mu$ (the average true similarity) overestimates the 
true distance. This is reasonable, since there is a prior that expects the true 
similarity to be around $\mu$.

The parameters $\mu$, $\sigma_{s}^2$, and $\sigma_{err,n}^2$ can be approximated 
using the method of moments over training data where there are enough common
users that $TrueSim$ can be approximated. This method requires approximating
$\sigma_{err,n}^2$ for every value of $n$ separately. To gain a more general 
model, we can model $\sigma_{err,n}^2$ as a function of $n$. Recall that 
$\sigma_{err,n}^2$ is the noisiness of the measurement $Sim$ about $TrueSim$ and $TrueSim = lim_{n \to
\infty}Sim$. Thus $Sim$ can be thought of as a sampling $n$ users from the
infinite set used to calculate $TrueSim$. Although $Sim$ may not be linear, we
the Central Limit Theorem motivates the intuition that the variance of $Sim$
would decrease as the $sqrt(n)$. Thus we model $\sigma_{err,n}^2$ as:
\begin{align}
\sigma_{err,n}^2 = \frac{\alpha}{\sqrt{n}},
\end{align}
yielding the single parameter model:
\begin{align}
\left(Y - \mu\right) = \frac{\sigma_{s}^2}{\sigma_{s}^2+\frac{\alpha}{\sqrt{n}}}
\left(X_n-\mu\right)
\end{align}
In terms of a modified similarity function, as desired earlier, we now have:
\begin{align}
ModSim(Sim, n) = \frac{\sigma_{s}^2}{\sigma_{s}^2+\frac{\alpha}{\sqrt{n}}}
\left(Sim-\mu\right) + \mu
\end{align}

\subsection*{Technique}
Based on this maodel, we try three different techniques to predict the true similarity from a
measured similarity. The techniques vary in their fidelity to the model. For
each, we limit our data to all pairs of items that have enough common users that
we can approximate $TrueSim \approx Sim$. Then we pick subsets of the common
users of various sizes $n$ and calculate $Sim_{n}$. With examples of $Sim$, $n$
and the resulting $TrueSim$, we can implement supervised learning.

@@

Using this model, we may view the problem of estimating the similarity $Y$ of
items $A$ and $B$ that most likely produced ratings $R_A$ and $_B$ as
that of finding the Maximum Likelihood Estimate of $Y | \vec{R}$:
\begin{align}
P(Y | \vec{R}) = \frac{P(\vec{R} | Y)P(Y)}{P(\vec{R})}.
\end{align}

\begin{align}
\hat{y} &= \argmax_yP(Y=y|\vec{R}=\vec{r}) 
\\&= \argmax_y\left[\frac{P(\vec{R}=\vec{r} |
Y=y)P(Y=y)}{P(\vec{R}=\vec{r})}\right]
\\&= \argmax_y\left[P(\vec{R}=\vec{r}|Y=y)P(Y=y)\right]
\\&=
\argmax_y\left[\frac{1}{2\pi\sqrt{\left|\mathbf{\Sigma_y}\right|}}\exp{\left(-\frac{1}{2}
(\vec{r} - \vec{\mu_r})^\mathrm{T}\mathbf{\Sigma_y}^{-1}(\vec{r} - \vec{\mu_r})\right)}
\frac{1}{\sigma_{s}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu_s)^2}
{2\sigma_{s}^2}\right)}\right]
\\&= \argmin_y\left[(\vec{r} -
\vec{\mu_r})^\mathrm{T}\mathbf{\Sigma_y}^{-1}(\vec{r} - \vec{\mu_r}) + \frac{(y -
\mu_s)^2}{\sigma_{s}^2} + \ln{\left|\mathbf{\Sigma_y}\right|} \right]
\\&=
\argmin_y\left[\vec{r^\prime}^\mathrm{T}\mathbf{\Sigma_y}^{-1}\vec{r^\prime}
 + \frac{(y - \mu_s)^2}{\sigma_{s}^2} + \ln{\left|\mathbf{\Sigma_y}\right|} \right]
\\&= \argmin_y\left[\frac{1}{\sigma_r^2}\frac{1}{1 -
y^2}\left[r_A^\prime(r_A^\prime - yr_B^\prime) + r_B^\prime(r_B^\prime -
yr_A^\prime)\right] + \frac{(y - \mu_s)^2}{\sigma_{s}^2} +
\ln{\left[\sigma_r^2(1 - y^2)\right]} \right],
\end{align}
where we used the fact that
\begin{align}
\mathbf{\Sigma_y}^{-1} = \frac{1}{\sigma_r^2}
\left[ \begin{array}{cc}  1 & y \\ y & 1 \end{array} \right]^{-1} =
\frac{1}{\sigma_r^2}\frac{1}{1 - y^2}\left[ \begin{array}{cc}  1 & -y \\ -y & 1
\end{array} \right].
\end{align}
and that
\begin{align}
\left|\mathbf{\Sigma}\right| = \sigma_r^2(1 - y^2)
\end{align}
along with the substitution $\vec{r^\prime} = (r_A^\prime~~r_B^\prime) = \vec{r}
- \vec{\mu_r}$. Next, we find the minimum by taking the derivative with respect
to $y$ and setting it to zero:
\begin{align}
\frac{d}{dy}\left[ \frac{1}{\sigma_r^2}\frac{1}{1 -
y^2}\left[r_A^\prime(r_A^\prime - yr_B^\prime) + r_B^\prime(r_B^\prime -
yr_A^\prime)\right] + \frac{(y - \mu_s)^2}{\sigma_{s}^2} +
\ln{\left[\sigma_r^2(1 - y^2)\right]} \right] &= 0 \\
-r_A^\prime r_B^\prime y^2 + ({r_A^\prime}^2 + {r_B^\prime}^2)y - r_A^\prime
r_B^\prime + \frac{\sigma_r^2}{\sigma_s^2}(1 - y^2)^2(y - \mu_s) - \sigma_r^2(1
- y^2)y &= 0.
\end{align}
This final expression implicitly yields the MLE of $y$ as a function of the
ratings $r_A$ and $r_B$. That is, constraining the parameters $\mu_r$,
$\sigma_r$, $\mu_s$, and $\sigma_s$ to realistic values, this expression defines
a surface that can be to efficiently estimate $y$ given a pair of ratings.

>>>>

@@

In our first approach, we model true and observed similarity probabilistically. The
goal is to use the observed similarity score directly to approximate the true similarity.
In this way, we can consider an abstact $TrueSim$ which ignores the details of 
$TrueSimRand$ or $TrueSimPref$. Similarly, we consider $Sim_{n}$ and $GoldSim$ to be
the generic observed similarity scores. Let $Y$ be a random variable representing 
the $TrueSim$ of a randomly chosen pair of items. Let $X_{n}$ be a random variable 
representing the $Sim_{n}$ of a pair of items when $n$ common users are observed.

@@

The model suggests that there is a linear correlation between $X_n$ ($Sim_n$)
and $Y$ ($TrueSim$). Moreover, the best predicted $TrueSim$ is a linear 
combination of the mean and the observed similiarity. Since $\sigma_{err,n}^2$ is 
nonnegative, the slope is always less than or equal to one, meaningthat the 
observed distance from $\mu$ (the average true similarity) overestimates the 
true distance. This is reasonable, since there is a prior that expects the true 
similarity to be around $\mu$.

This model requires calculating $\sigma_{err,n}^2$ for every value of $n$ 
separately. To gain a more general model, which relates the $\sigma_{err, n}$'s, 
we model $\sigma_{err,n}^2$ as a function of $n$. Recall that $\sigma_{err,n}^2$ 
is the noisiness of the observed $Sim_n$ about $TrueSim$. Since the $TrueSim$ is
the $Sim$ when all users that will rate the items have rated them, $Sim_n$ can be
thought of as sampling $n$ users from the set used to calculate $TrueSim$. 
Although $Sim$ is likely more complicated than an average, the Central Limit 
Theorem motivates the intuition that the variance of $Sim$ would decrease as the 
$\sqrt{n}$. Thus we model $\sigma_{err,n}^2$ as:
\begin{align}
\sigma_{err,n}^2 = \frac{\alpha}{\sqrt{n}},
\end{align}
yielding the single parameter model:
\begin{align}
\left(Y - \mu\right) = \frac{\sigma_{s}^2}{\sigma_{s}^2+\frac{\alpha}{\sqrt{n}}}
\left(X_n-\mu\right)
\end{align}

\subsection*{Technique}
The parameters $\mu$, $\sigma_{s}^2$, and $\sigma_{err,n}^2$ can be approximated 
using the method of moments over training data. $\mu$ and $\sigma_{s}$ are
calculated

Based on this maodel, we try three different techniques to predict the true similarity from a
measured similarity. The techniques vary in their fidelity to the model. For
each, we limit our data to all pairs of items that have enough common users that
we can approximate $TrueSim \approx Sim$. Then we pick subsets of the common
users of various sizes $n$ and calculate $Sim_{n}$. With examples of $Sim$, $n$
and the resulting $TrueSim$, we can implement supervised learning.

@@

Using this model, we may view the problem of estimating the similarity $Y$ of
items $A$ and $B$ that most likely produced ratings $R_A$ and $_B$ as
that of finding the Maximum Likelihood Estimate of $Y | \vec{R}$:
\begin{align}
\hat{y} &= \argmax_yP(Y=y|\vec{R}=\vec{r}) 
\\&= \argmax_y\left[\frac{P(\vec{R}=\vec{r} |
Y=y)P(Y=y)}{P(\vec{R}=\vec{r})}\right]
\\&= \argmax_y\left[P(\vec{R}=\vec{r}|Y=y)P(Y=y)\right]
\\&=
\argmax_y\left[\frac{1}{2\pi\sqrt{\left|\mathbf{\Sigma_y}\right|}}\exp{\left(-\frac{1}{2}
(\vec{r} - \vec{\mu_r})^\mathrm{T}\mathbf{\Sigma_y}^{-1}(\vec{r} - \vec{\mu_r})\right)}
\frac{1}{\sigma_{s}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu_s)^2}
{2\sigma_{s}^2}\right)}\right]
\\&= \argmin_y\left[(\vec{r} -
\vec{\mu_r})^\mathrm{T}\mathbf{\Sigma_y}^{-1}(\vec{r} - \vec{\mu_r}) + \frac{(y -
\mu_s)^2}{\sigma_{s}^2} + \ln{\left|\mathbf{\Sigma_y}\right|} \right]
\\&=
\argmin_y\left[\vec{r^\prime}^\mathrm{T}\mathbf{\Sigma_y}^{-1}\vec{r^\prime}
 + \frac{(y - \mu_s)^2}{\sigma_{s}^2} + \ln{\left|\mathbf{\Sigma_y}\right|} \right]
\\&= \argmin_y\left[\frac{1}{\sigma_r^2}\frac{1}{1 -
y^2}\left[r_A^\prime(r_A^\prime - yr_B^\prime) + r_B^\prime(r_B^\prime -
yr_A^\prime)\right] + \frac{(y - \mu_s)^2}{\sigma_{s}^2} +
\ln{\left[\sigma_r^2(1 - y^2)\right]} \right],
\end{align}
where we used the fact that
\begin{align}
\mathbf{\Sigma_y}^{-1} = \frac{1}{\sigma_r^2}
\left[ \begin{array}{cc}  1 & y \\ y & 1 \end{array} \right]^{-1} =
\frac{1}{\sigma_r^2}\frac{1}{1 - y^2}\left[ \begin{array}{cc}  1 & -y \\ -y & 1
\end{array} \right].
\end{align}
and that
\begin{align}
\left|\mathbf{\Sigma}\right| = \sigma_r^2(1 - y^2)
\end{align}
along with the substitution $\vec{r^\prime} = (r_A^\prime~~r_B^\prime) = \vec{r}
- \vec{\mu_r}$. Next, we find the minimum by taking the derivative with respect
to $y$ and setting it to zero:
\begin{align}
\frac{d}{dy}\left[ \frac{1}{\sigma_r^2}\frac{1}{1 -
y^2}\left[r_A^\prime(r_A^\prime - yr_B^\prime) + r_B^\prime(r_B^\prime -
yr_A^\prime)\right] + \frac{(y - \mu_s)^2}{\sigma_{s}^2} +
\ln{\left[\sigma_r^2(1 - y^2)\right]} \right] &= 0 \\
-r_A^\prime r_B^\prime y^2 + ({r_A^\prime}^2 + {r_B^\prime}^2)y - r_A^\prime
r_B^\prime + \frac{\sigma_r^2}{\sigma_s^2}(1 - y^2)^2(y - \mu_s) - \sigma_r^2(1
- y^2)y &= 0.
\end{align}
This final expression implicitly yields the MLE of $y$ as a function of the
ratings $r_A$ and $r_B$. That is, by constraining the parameters $\mu_r$,
$\sigma_r$, $\mu_s$, and $\sigma_s$ to realistic values, the equation
can be used to efficiently estimate $y$ given a pair of ratings.
